---
title: "enrichmeta Tutorial: Comprehensive Metabolite Enrichment Analysis with KEGG Pathways"
author: "Botao Zhang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{enrichmeta Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `enrichmeta` package provides a comprehensive toolkit for metabolite enrichment analysis using KEGG pathways. This tutorial demonstrates all core functionality:

1. Retrieving KEGG pathway data with caching
2. Performing enrichment analysis
3. Visualizing results
4. Pathway-metabolite mapping
5. Pathway name annotation
6. Analyzing shared metabolites between pathways

## Installation

```{r install, eval = FALSE}
# Install from GitHub
devtools::install_github("maxgodson1/enrichmeta")
```

## Load Package

```{r library}
library(enrichmeta)
```

## Step 1: Get KEGG Pathway Data

Use `getkeggdata()` to retrieve pathway data for a species. Data is cached locally for future use.

```{r get_data}
# Get data for zebrafish (Danio rerio)
kegg_data <- getkeggdata("dre", cache_dir = "./kegg_cache")

# Explore data structure
str(kegg_data, max.level = 1)

# Number of pathways and compounds
cat("Pathways:", length(kegg_data$pathways), "\n")
cat("Compounds:", length(kegg_data$compounds), "\n")
```

Data structure:
- `pathways`: Named list of pathway names (ID -> name)
- `path_cpd_map`: List of metabolites per pathway (ID -> vector of compounds)
- `compounds`: Named list of compound names (ID -> name)

## Step 2: Enrichment Analysis

Perform enrichment analysis with `enrichkegg()` using a set of metabolite KEGG IDs.

```{r enrichment}
# Example metabolite IDs (significant metabolites from your study)
sig_metabolites <- c("C00022", "C00024", "C00031", "C00033", "C00036",
                    "C00041", "C00042", "C00068", "C00074", "C00149")

# Run enrichment analysis
enrich_results <- enrichkegg(
  keggid = sig_metabolites,
  pathway_data = kegg_data
)

# View top results
head(enrich_results, 5)
```

## Step 3: Visualize Enrichment Results

### Bar Plot

```{r bar_plot, fig.width=8, fig.height=6}
plotenrichbar(
  results = enrich_results,
  top = 10,
  title = "Top Enriched Pathways in Zebrafish",
  low_color = "#e41a1c",  # Red for significant
  high_color = "#ffffcc" # Light yellow for non-significant
)
```

### Dot Plot

```{r dot_plot, fig.width=8, fig.height=6}
plotenrichdot(
  results = enrich_results,
  top = 10,
  title = "Enriched Pathways in Zebrafish",
  low_color = "#e41a1c",
  point_size_range = c(4, 10)
)
```

## Step 4: Pathway-Metabolite Mapping

### Find Pathways for Metabolites

```{r find_paths}
# Get pathways containing specific metabolites
metab_paths <- findcpdspaths(c("C00022", "C00031"), pathway_data = kegg_data)
head(metab_paths)
```

### Find Metabolites in Pathways

```{r find_cpds}
# Get metabolites in specific pathways
path_cpds <- findpathscpds(c("dre00010", "dre00020"), pathway_data = kegg_data)
head(path_cpds)
```

## Step 5: Annotate Pathway and Metabolite Names

### Map Pathway IDs to Names

```{r pathway_names}
# Get top 10 enriched pathway IDs
top_paths <- head(enrich_results$pathwayID, 10)

# Map to pathway names
path_names <- findpathsnames(top_paths, pathway_data = kegg_data)
path_names
```

### Map Metabolite IDs to Names

```{r compound_names}
# Extract metabolite IDs from enrichment results
all_metabs <- unique(unlist(strsplit(enrich_results$keggID, "; ")))

# Map to compound names
compound_names <- findcpdsnames(all_metabs, pathway_data = kegg_data)
head(compound_names)
```

## Step 6: Shared Metabolites Between Pathways

Analyze metabolite sharing between pathways:

```{r shared_cpds}
# Get top 5 enriched pathway IDs
top_paths <- head(enrich_results$pathwayID, 5)

# Find shared metabolites
shared_df <- findsharedcpds(
  pathwayid = top_paths,
  pathway_data = kegg_data,
  min_shared = 2  # Only show pairs with ≥2 shared metabolites
)

# View raw results
shared_df

# Enhance results with pathway names
if (nrow(shared_df) > 0) {
  # Create full pathway name mapping
  all_paths <- unique(c(shared_df$from, shared_df$to))
  path_name_map <- findpathsnames(all_paths, kegg_data)
  names(path_name_map)[2] <- "PathwayName"
  
  # Merge pathway names
  shared_df <- merge(shared_df, path_name_map, by.x = "from", by.y = "pathwayID")
  names(shared_df)[ncol(shared_df)] <- "from_name"
  
  shared_df <- merge(shared_df, path_name_map, by.x = "to", by.y = "pathwayID")
  names(shared_df)[ncol(shared_df)] <- "to_name"
  
  # Reorder columns
  shared_df <- shared_df[, c("from", "from_name", "to", "to_name",
                          "shared_count", "keggID")]
  
  # View enhanced results
  head(shared_df)
}
```

## Step 7: Complete Analysis Workflow

Combine all steps into a comprehensive analysis:

```{r workflow, eval = FALSE}
# 1. Get KEGG data
kegg_data <- getkeggdata("hsa", cache_dir = "kegg_cache")

# 2. Run enrichment analysis
enrich_results <- enrichkegg(sig_metabolites, kegg_data)

# 3. Visualize results
plotenrichbar(enrich_results, top = 15)
plotenrichdot(enrich_results, top = 15)

# 4. Get pathway names
top_paths <- head(enrich_results$pathwayID, 10)
path_names <- findpathsnames(top_paths, kegg_data)

# 5. Analyze shared metabolites
shared_df <- findsharedcpds(top_paths, kegg_data, min_shared = 3)

# 6. Map metabolite names
if (nrow(shared_df) > 0) {
  shared_metabs <- unique(unlist(strsplit(shared_df$keggID, ";")))
  metab_names <- findcpdsnames(shared_metabs, kegg_data)
}
```

## Conclusion

This tutorial covered the full functionality of `enrichmeta`. Key features demonstrated:

1. **Data Acquisition**: `getkeggdata()` for pathway data retrieval and caching
2. **Enrichment Analysis**: `enrichkegg()` for metabolite set enrichment
3. **Visualization**: `plotenrichbar()` and `plotenrichdot()` for result exploration
4. **Mapping Tools**:
   - `findcpdspaths()`: Find pathways containing metabolites
   - `findpathscpds()`: Find metabolites in pathways
   - `findpathsnames()`: Map pathway IDs to descriptive names
   - `findcpdsnames()`: Map metabolite IDs to compound names
5. **Pathway Relationships**: `findsharedcpds()` for analyzing shared metabolites

For advanced usage and detailed function parameters:
```{r help, eval=FALSE}
help(package = "enrichmeta")
```

## References

1. Kanehisa, M., & Goto, S. (2000). KEGG: Kyoto Encyclopedia of Genes and Genomes. Nucleic Acids Research, 28(1), 27-30.
